### 2017-07-06 inference.tex
Translation to MSFOL.
Sec. 4 Part 1.

@@

>>>>

Rely-Guarantee framework presented in the previous section facilitates
modular proofs for weakly-isolated transactions, but imposes a
non-trivial annotation burden on the programmer. In particular, it
requires each statement ($c$) of the transaction to be annotated with
a stable pre- ($P$) and post-condition ($Q$), and loops to be
annotated with stable inductive invariants ($\psi$). While weakest
pre-condition style predicate transformers can help in inferring $P$'s
and $Q$'s for regular statements, inferring inductive loop invariants
is still a challenge, even for the simpler form of loops we admit.
Thus, predicate transformer-based approach to proof automation is
unlikely to be effective in this case. We therefore present an
alternative approach to RG proof automation in this section that is
based on state transformers instead of predicate transformers. The
idea is to infer the effect that each statement has on the
transaction-local database state $\stl$ (i.e., how it transforms
$\stl$), and compose multiple such effects together to describe the
effect of the transaction as a whole. Importantly, this approach
generalizes to loops, where the effect of a loop can be inferred as a
well-defined function of the effect of its body, thanks to certain
pleasant properties enjoyed by the database programs modeled by
$\txnimp$.

\input{logic-syntax}

At the core of our approach is a simple language ($\SL$) to express
set transformations shown in Fig.~\ref{fig:logic-syntax}. The language
admits only set expressions that include variables ($x$), literals of
the form $\{x \,|\, \phi^0\}$ where $\phi^0$ is a propositional
(quantifier-free) formula on $x$, a restricted form of existential
quantification that binds a set $x$ satisfying proposition $\phi^0$ in
a set expression $s$, a monadic composition of two set expressions
($s_1$ and $s_2$) using bind ($\bind$), a conditional set expression
where the condition is a propositional formula, and a union of two set
expressions. Symbols $\stl$ and $\stg$ are also variables in $\SL$,
but are used to denote local and database states, respectively.  The
language is carefully chosen to be expressive enough to capture the
semantics of $\txnimp$ statements, yet simple enough to have a
semantics-preserving translation to the first-order logic.

@@

Fig.~\ref{fig:inference-rules} shows the syntax-directed inference
rules for $\txnimp$ commands inside a transaction $\C{TXN}_i$.
Inference depends on the $\I$-constrained rely relation $\R$, and the
high-level invariant $I$ for the reasons that will become clear
shortly.  Inference rules compute, for each command $c$, a (meta)
function $\F$ that returns a set as an expression in $\SL$, given a
pair of sets ($\stl$ and $\stg$) that describe local and a global
database states, respectively. The expression returned by
$\F(\stl,\stg)$ (abstractly) describes the set of records that get
added to $\stl$ as a result of executing $c$ under $\stl$ and $\stg$.
Thus, $\F$ captures the \emph{effect} part of the state transformer of
$c$, which is the function $\lambda(\stl,\stg).~\stl \cup
\F(\stl,\stg)$. For $\F$ to be useful in RG verification, it needs to
be stable w.r.t the rely relation $\R$. The stability condition on
effects is defined below:
\begin{smathpar}
\begin{array}{lcl}
  \stable(\R,\F) & \Leftrightarrow & \forall \stl,\stg,\stg'.~
  \R(\stl,\stg,\stg') \Rightarrow \F(\stl,\stg) \equiv \F(\stl,\stg')
\end{array}
\end{smathpar}
Intuitively, the stability condition requires an effect to describe
the same set of records before and after the interference. The
equivalence in $\SL$ translates to the equality in first-order logic,
as will be explained shortly. In the inference rules, stability is
enforced constructively by a meta function $\stabilize{\cdot}$, which
accepts an effect and returns a new effect that is guaranteed to be
stable under $\R$. $\stabilize{\cdot}$ achieves the stability
guarantee by abstracting away the bound $\stg$ in an unstable $\F$ to
an existentially bound $\stg'$ as described below:
\begin{smathpar}
\begin{array}{lcll}
  \stabilize{\F} & = & \F & \texttt{if } \stable(\R,\F)\\
  & = & \lambda (\stl,\stg).~\existsl(\stg',I(\stg'),\F(\stl,\stg')) & \texttt{otherwise}\\
\end{array}
\end{smathpar}
Observe that when $\F$ is not stable, then $\stabilize{\F}$ returns an
$\F'$ that simply ignores its $\stg$ argument in favor of a generic
$\stg'$, making $\F'$ trivially stable. The existentially bound
$\stg'$ satisfies $I$ since commands always executed against a valid
database state.

The inference rules of SQL statements, like the RG rules, closely
follow the corresponding reduction rules in Fig.~\ref{fig:txnimp},
except that their language of expression is $\SL$. For instance, while
the reduction rule for \C{UPDATE} declaratively specifies the set of
updated records, the inference rule uses $\SL$'s bind operation to
\emph{compute} the set. Other SQL rules do likewise. The inference
rules for \C{LET} binders, conditionals, and sequences compose the
effects inferred for their subcommands. For instance, the effect of a
sequence of commands $c_1;c_2$ is the union of effects $F_1$ and $F_2$
of $c_1$ and $c_2$, respectively, except that $F_2$ is applied to a
state ($\stl$) to which $F_1$ has already been applied, reflecting
their order of reduction. The inference rule for \C{FOREACH} takes
advantage of the $\SL$'s bind operator to lift the effect inferred for
the loop body to the level of the loop. Since records added to $\stl$
in each iteration of \C{FOREACH} are independent of the previous
iteration, sequential composition of the effects of different
iterations is same as their parallel composition. Since the loop body
is executed once per each $z\in x$, the effect of the the loop is the
union of effects ($\F$) for all $z\in x$, all applied to the same
state ($\stl$ and $\stg$). That is, $\F_{loop}(\stl,\stg) =
\bigcup_{z\in x}\F_{body}(\stl,\stg)$. From the definition of
set monad's bind, $\F_{loop}(\stl,\stg) = x \bind (\lambda
z.~F_{body}(\stl,\stg))$, which is what the rule infers.

\input{logic}

\begin{theorem}
if $c \elabsto \F$ and $\semof{\F(\stl,\stg)} = (\phi^1,g)$ and
$P(\stl,\stg) = true$ and $Q(\stl',\stg) = \exists\stl.~ \phi^1 \wedge
  \forall a. a\in\stl' \Leftrightarrow a\in\stl \vee g(a)$, then
  $\R \vdash \hoare{P}{c}{Q}$
\end{theorem}
